---
title: "Generate Model Input Data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate Model Input Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

```{r libraries}

library(spopmodel)

```

## Background

A key step in `spopmodel` employs a female-based Leslie Matrix. Thus, this model requires data about survival probability, reproduction (spawning probability & fecundity), and age distribution (frequency). Here we show --- using raw age & length data --- the processes by which we generate model inputs. Ultimately, we use these inputs (or starting data) to run simulations (see vignette of same name) that then "feed" the model.

`spopmodel` was developed to assess San Francisco-estuary based population dynamics. Herein we use the data on which the model was built (i.e., White Sturgeon data collected by the California Department of Fish & Wildlife [CDFW] in collaboration with the US Fish & Wildlife Service and the University of Idaho [the model developers]).

*Note*: about data not exact to that used in thesis
*Note*: if you already have these data proceed to sims
*Note*: ??

## Data

For this demonstration, we'll use dataset `trammel_catch`. These data are from the CDFW's [Sturgeon Study](https://www.wildlife.ca.gov/Conservation/Delta/Sturgeon-Study). This particular dataset is from 2014-2016 sampling. It contains age & length data, with `r sum(!is.na(trammel_catch[["Age"]]))` aged fish from `r nrow(trammel_catch)` observations. Age was estimated from fin ray samples collected in the field.

*Note*: Herein we **do not** adjust catch data for gear selectivity. Likely, such an adjustment would increase catch in most of the length bins (and thus ages). A vignette about performing such adjustment is forthcoming.

Let's plot FL ~ Age to get some sense of range and mean-length-at-age. We added some noise (`jitter()`) to variable Age to reduce over-plotting.

```{r plot-raw}

plot(FL ~ jitter(Age), data = trammel_catch, col = "steelblue")
abline(h = 190, col = "grey70", lty = 2)

```

We arbitrarily set a horizontal line at 190 cm FL and noted we do not have many *aged* fish &ge; 190. If fact, as we see below, the entire dataset only has `r nrow(trammel_catch[trammel_catch[["FL"]] > 190, ])` fish &ge; 190 cm FL. Thus, to ensure no empty bins in our age-length key, we will (somewhat arbitrarily) set age to 19 for "age-less" fish &ge; 190 cm FL.

*Note*: In reality, we may decide to age these large fish if we had the means to do so (i.e., had a fin-ray sample). But for this purpose, age-19 is a pretty good surrogate give aged fish of similar size. We also could manipulate bin size (in length frequency) to see if these fish are binned with any aged fish.

```{r fish-gte190}

bool_gte190 <- trammel_catch[["FL"]] >= 190

trammel_catch[bool_gte190, ]

```

We can replace `NA` with 19 simply using `is.na()` along with our Boolean above. We display the results to verify our change.

```{r age-19}

bool_ageNA <- is.na(trammel_catch[["Age"]])

trammel_catch[bool_gte190 & bool_ageNA, "Age"] <- 19

# verification
trammel_catch[bool_gte190, ]

```

## Model Inputs

`spopmodel`'s design requires four data inputs: age distribution; spawning probability; egg count; and survival probability. All four must configure to the age range established in age distribution, which should contain young age classes (i.e., ages-0 to ages-2). These datasets are not fed directly to the model but rather are used to run stochastic simulations. The model directly uses these simulations to generate population growth rate ($\lambda$) predictions.

Though not directly a model input, we begin by creating a length frequency distribution. As we'll see in the next section, length frequency if useful for --- among other things --- aging the un-aged portion.   

### Length Frequency

To age the un-aged portion of our data, we have a couple of methods from which to choose. Both methods --- using age-length key or assigning age to each fish --- require length frequency. We use `Frequency()` for such a task, setting bin width to 5 cm. (For White Sturgeon, 5 cm is a good starting bin size, but you may want to experiment with various sizes. Here, we will stick with 5 cm.) As shown below, `Frequency()` output provides a list with nine elements.

```{r len-freq}

len_freq <- Frequency(trammel_catch[["FL"]], binWidth = 5)

str(len_freq)

```

For curiosity, we'll run some descriptive stats on variable FL. We see a range from `r len_freq$xstats()$Min` to `r len_freq$xstats()$Max` cm FL.

```{r len-stats}

len_desc_stats <- len_freq$xstats()

# for display
unlist(len_desc_stats)

```

We also note from `len_freq` we get breaks (or bins; see `len_freq$breaks`). We'll use this in the next step (below) to age each fish. Plotting length-frequency we see our data is skewed right. The red vertical line denotes the median (`r len_desc_stats$Med` cm FL).

```{r plot-lf}

plot(len_freq)

```

```{r make-alk, include=FALSE, eval=FALSE}

# from narrative not used: create our age-length key. `sum(is.nan(alk))` checks
# for empty bins and value should be 0 (i.e., no empty bins).

alk <- MakeALKey(
  data = trammel_catch,
  len = FL,
  age = Age,
  lenBreaks = len_freq$breaks
)

sum(is.nan(alk))

```

### Age Distribution

`AgeEach()` is modeled after `FSA::alkIndivAge()`. The stripped-down output (i.e., just the ages) of `AgeEach()` is better suited for `spopmodel`. (**Note** however more testing of `AgeEach()` is still required as of 15-Oct-2018.)

```{r age-each}

ages <- AgeEach(
  data = trammel_catch,
  len = FL,
  age = Age,
  lenBreaks = len_freq$breaks
)

```

```{r compare-FSA, include=FALSE, eval=FALSE}

# TODO: correct flaw in logic with AgeEach(); is not assign ages correctly
#       a good check is to plot fl ~ age, should see increasing mean with age
#       (J. DuBois 12-Oct-2018)

# 15-Oct-2018: corrected flaw (I feel) using unsplit instead of unlist in
# AgeEach(). Correction works and CheckAge() appears to age quite similarly to
# FSA::alkIndivAge --- check with code below, though not identical due to use of
# sample() --- would be wise to do more checking just to ensure proper
# functioning of CheckAge()

# fsa_indiv <- FSA::alkIndivAge(
#   key = alk,
#   formula = ~FL,
#   data = trammel_catch,
#   breaks = len_freq$breaks
# )
# 
# plot(sort(fsa_indiv$age), y = sort(ages))
# abline(a = 0, b = 1, col = 2)
# 
# table(fsa_indiv$age)-
# table(ages)

```

Now we can simply call R's `table()` to generate age-frequency. We now have our age-frequency (for ages 3-19). (Sum of `age_freq` should equal 1000. You can double-check if you like.)

```{r age-freq}

age_freq <- table(ages[["Ages"]], dnn = NULL)

age_freq

```

```{r mean-laa}

# mean length at each age (3-19)
mean_len_age <- aggregate(
  trammel_catch[["FL"]],
  by = list(Age = ages[["Ages"]]), 
  FUN = mean
)

# something more descriptive than 'FL'
colnames(mean_len_age)[2] <- "MeanFL"

mean_len_age

```

Let's plot our raw length ~ age data, overlayed with mean-length-at-age (red '+' sign). It looks reasonable and as we expected.

```{r plot-mean_laa}

plot(jitter(ages[["Ages"]]), trammel_catch[["FL"]], col = "grey70", las = 1)
points(mean_len_age, pch = "+", col = "darkred")

```

We now need an age distribution based on a starting population. We use `AgeDist()` passing our `age_freq` variable and keeping defaults `abund = 48000` and `fracFemale = 0.5`. The output (as shown below) is the basis for how data `age_dist` was created. (With `spopmodel` library loaded, view `age_dist` & compare to `age_distribution`.)

```{r age-dist}

age_distribution <- AgeDist(ageFreq = age_freq)

age_distribution

```

Though we have our age-distribution, we still need values for ages 0-2. Per model development, age-1 & -2 values were obtained using linear regression (specifically log-linear).

For convenience, we assign estimated age abundance (`EstAgeAbund`) to variable `est_abundance`. Estimated age abundance is simply age-frequency (as a proportion of total) multiplied by our starting abundance (in this case `r age_distribution$OverallAbund`). We assign our x & y values accordingly, use R's `lm()` for linear regression, and then plot the results. As we expect, frequency declines with age.

```{r age1-2}

est_abundance <- age_distribution[["EstAgeAbund"]]

# variables for plotting & linear regression
yval <- log(as.vector(est_abundance))
xval <- as.numeric(names(est_abundance))

mod <- lm(yval ~ xval)

plot(
  x = xval,
  y = yval,
  type = "b",
  xlab = "Age",
  ylab = "log(EstAbund)",
  panel.first = abline(mod, col = "grey60", lty = 2),
  # panel.first = grid(),
  las = 1
)

```

We use our linear model to predict ages 1 & 2. These values, however, include males & females. We need only females and will combine ages 1 & 2 with ages 3-19 (females only).

```{r age1_2-pred}

age1_2 <- exp(predict(object = mod, newdata = list(xval = c(1, 2))))

females <- c(
  age1_2 * age_distribution$FracFemale,
  age_distribution$CountFemByAge
)

females

```

### Spawning Probability

Later we will calculate age-0 distribution (frequency) and append `females` accordingly. To do so we need spawning probability, which we will create now.

For our White Sturgeon population, we do not have age-specific spawning probabilities. Thus, we rely on our age & length data plus probability data from previous research.

Running `SpawningProb()` will issue a warning saying "Using `SpawningProbWST` for `pMat`". `pMat` is short for probability of maturity and in lieu of our own dataset, we use data from Champman (1989), termed herein `SpawningProbWST`. These data can be found [here](https://github.com/jasondubois/spopmodel/blob/master/data-raw/spawning.csv).

Internally, `SpawningProb()` performs `glm()` (as probability ~ length) using `SpawningProbWST`, and then from this model predicts probability given mean-length-at-age. The resulting `Prob` in `p_spawn` is the predicted probability multiplied by the argument supplied to `mature` (in this case default 0.15, or 15% of all females spawn annually). `Err` is probability multiplied by 0.20.

*Note*: `glm()` inside `SpawningProb()` may issue the following warning:  non-integer #successes in a binomial glm! As of 24-Oct-2018, we are still looking into this.

```{r prob-spawn}

bool_mlaa_gt9 <- mean_len_age$Age > 9


# using default argument for mature parameter (0.15)
p_spawn <- with(data = mean_len_age[bool_mlaa_gt9, ], expr = {
  SpawningProb(len = MeanFL, age = Age)
})

p_spawn

```

We've subsetted on fish "> age-9" because --- for White Sturgeon --- females mature beginning around age-10. So to complete our dataset (i.e., probability of spawning) we need to `rbind()` ages 0-9.

We'll preserve `p_spawn` for future use and name our new dataframe `prob_spawn2` so we don't confuse with `prob_spawn` internal to `spopmodel`. In fact, it's not a bad idea to compare `prob_spawn` with `prob_spawn2`, as both datasets should be roughly the same.

```{r prob-spawn_full}

prob_spawn2 <- rbind(
  data.frame(Age = 0:9, Prob = 0, Err = 0),
  p_spawn[, c("Age", "Prob", "Err")]
)

prob_spawn2

```

### Age Distribution: age-0

Now we return to age distribution. Estimating age-0 frequency is a bit more involved. We need first to calculate mean length at age, given our age and length data.

Here we employ Devore's (et al. 1995) equation to calculate number of eggs based on fork length ($eggs=0.072 \times l_i^{2.94}$, where *l~i~* is mean length at age. Like maturity, we subset on fish > age-9.

```{r devore}

eggs_female <- 0.072 * (mean_len_age[bool_mlaa_gt9, "MeanFL"]^2.94)

```

Now that we have number of eggs per spawning female, we need to multiply this by the number of females and spawning probability.

```{r age0-freq}

age0 <- sum(p_spawn[["Prob"]] * females[10:19] * eggs_female)

```

We get a value of `r format(age0, big.mark = ",")` age0 fish. We will append this to our `females` variable. Our age distribution really should be a dataframe, so we can construct that now. We name this dataframe `age_dist2` so as not to confuse it with `age_dist` internal to `spopmodel`. (The two dataframes should be roughly the same, although you likely will find a big (and yet unexplained) difference between age-0 values.)

```{r age_dist2}

age_dist2 <- data.frame(
  Age = as.numeric(c(0, names(females))),
  Freq = c(age0, unname(females)),
  row.names = NULL
)

age_dist2

```

### Egg Count

We need age-specific egg count. Absent this, the process to generate such data is similar to the one that creates spawning probability. Here we use `EggCount()`, which runs linear regression on Devore (et al. 1995) [fecundity data](https://github.com/jasondubois/spopmodel/blob/master/data-raw/fecundity.csv) (here in named `FecundityWST`) and then predicts fecundity given mean-length-at-age. 

Age, we `rbind()` ages-0 through -9 data as 0, given maturity begins around age-10. We assign this to variable `num_eggs` so as not to confused with `numbers_eggs` internal to `spopmodel`.

*Note*: the use of linear regression is peculiar given Devore's egg count equation (above). We are following model protocol, but you may want to experiment by plugging mean-length-at-age directly into Devore's equation. 

```{r num-eggs}

# egg count age-10 to age-19
egg_count <- with(data = mean_len_age[bool_mlaa_gt9, ], expr = {
  EggCount(len = MeanFL, age = Age)
})

num_eggs <- rbind(
  data.frame(Age = 0:9, Count = 0, Err = 0),
  egg_count[, c("Age", "Count", "Err")]
)

num_eggs

```

### Survival Probability

Here we create the final of four data inputs: survival probability. Because exploitation ($\mu$) affects overall survival, we need to derive survival probability for every level of $\mu$. We do this using `SurvivalProb()`.

`SurvivalProb()` requires age-specific probability survival, along with some measure of error. Here we use data from `prob_survival` (see `help(prob_survival)`). For ages subject to harvest (i.e., legal-sized fish), `SurvivalProb()` adjusts survival rate --- i.e., in this case `max(prob_survival[1:4, "prob"])` --- given fishing mortality (*F*), as calculated below. *Note*: `max(prob_survival[1:4, "prob"])` (or `r max(prob_survival[1:4, "prob"])`) assumes no fishing mortality.

$$F=\frac{\mu \times Z}{A}$$

where
$Z=-log(S)$, S is `estS` (see `help(prob_survival)`)  
$\mu$, is `estMu` (see `help(prob_survival)`)  
$A=1-S$

**Note**: the derivation for standard error of fish affected by harvest is not well understood (by me - J. DuBois, that is) at this point. We simply multiply `sRateErr` by 0.45. This will suffice for now until we better understand the method used to derive this SE. I used 0.45 because the values in `prob_survival` are 0.04281 for non-harvestable fish &ge; age-3 and 0.01932 for legal-sized fish, 0.04281 * 0.45 is roughly 0.019.

**Note**: we use values for ages 0-2 culled from literature. See S. Blackburn's Masters Thesis for references or `help(prob_survival)`.

`SurvivalProb()` accepts a numeric vector as an argument for parameter `mu`. In fact, it's very likely you'll want to pass a vector with varying `mu`s for use in simulations. Here we create a very small vector (3 levels of `mu`) just for ease of demonstration. In reality, this model was developed on `mu` 0 to 0.30, by 0.01. For multiple `mu`s, `SurvivalProb()` will return a named list using "mu_" followed by the level of `mu`.

```{r prob-surv}

mus <- c(0.01, 0.02, 0.03)

prob_survival2 <- SurvivalProb(
  ages = prob_survival[["age"]],
  sRate = prob_survival[1:4, "prob"],
  sRateErr = prob_survival[1:4, "se"],
  mu = mus,
  agesMu = 10:15
)

str(prob_survival2)

```

Below is a snippet of dataframe `mu_0.01` to show ages not affected by harvest (7-9) and those affected by harvest (10-12).

*Note*: Per model design (and mostly due to data gaps) we use (1) constant survival rate for ages 3-19, less of course fishing mortality and (2) constant standard error throughout, irrespective of mu. It might be an interesting exercise to vary age-specific survival, perhaps at least for older fish.

```{r prob-surv_display}

# just a snippet for display
prob_survival2$mu_0.01[8:13, ]

```

## References

LEFT OFF HERE 24-Oct-2018
